---
title: "EDA"
author: "Mustapha Mustapha"
date: "2024-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## My sleep data - part 2 Exploratory Data Analysis
Prompt: What is exploratory data analysis?
Exploratory data analysis (EDA) is an approach to analyzing data sets to summarize their main characteristics, often with visual methods. A statistical model can be used or not, but primarily EDA is for seeing what the data can tell us beyond the formal modeling or hypothesis testing task.

# Steps in EDA
1. Data Collection (done) - collected and imported sleep data from Fitbit and calculated sleep_hours - sum of daily sleep in hours
2. Data Cleaning
3. Data Preprocessing
4. Data Exploration
5. Data Visualization
6. Data Modeling

# Step 2: Data Cleaning
Data cleaning is the process of detecting and correcting (or removing) corrupt or inaccurate records from a record set, table, or database and refers to identifying incomplete, incorrect, inaccurate or irrelevant parts of the data and then replacing, modifying, or deleting the dirty or coarse data.

In this step, we will: detect outliers in sleep_hours variable

```{r pressure, TRUE}
#View(sleep_daily)
str(sleep_daily)
summary(sleep_daily)
# sleep_daily$sleep_hours : num [1:721] 6.67 6.93 10.45 3.73 6.73 ...
# remove outliers from sleep_hours >15 or sleep_hours < 3

sleep_daily_clean <- sleep_daily[sleep_daily$sleep_hours < 16 & sleep_daily$sleep_hours > 2,]

```
### **Exploratory Analyses:**
1. **Summary Statistics:** Calculate the mean, median, standard deviation, and range of daily sleep duration to understand the central tendency and variability.
```{r summary_stats, TRUE}
print(paste("Mean sleep hours (with outliers): ", round(mean(sleep_daily$sleep_hours), 2)))
print(paste("Mean sleep hours (no outliers): ", round(mean(sleep_daily_clean$sleep_hours), 2)))
print(paste("Median sleep hours (with outliers): ", round(median(sleep_daily$sleep_hours), 2)))
print(paste("Median sleep hours (no outliers): ", round(median(sleep_daily_clean$sleep_hours), 2)))
print(paste("SD of sleep hours (with outliers): ", round(sd(sleep_daily$sleep_hours), 2)))
print(paste("SD of sleep hours (no outliers): ", round(sd(sleep_daily_clean$sleep_hours), 2)))
print(paste("Range of sleep hours (with outliers): ", round(min(sleep_daily$sleep_hours), 2), 
      ", ", round(max(sleep_daily$sleep_hours), 2)))
print(paste("Range of sleep hours (no outliers): ", round(min(sleep_daily_clean$sleep_hours), 2),
      ", ", round(max(sleep_daily_clean$sleep_hours), 2)))


```

## 2. **Sleep Distribution:** 
Plot the distribution of daily sleep duration (e.g., histogram or kernel density estimate) to visualize the overall pattern.
```{r sleep_distribution, TRUE}
ggplot(sleep_daily_clean, aes(x = sleep_hours)) +
  geom_histogram(aes(y = ..density..), binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_density() +
  geom_vline(aes(xintercept = median_sleep), color = "red", linetype = "dashed", size = 1) +
  labs(x = "Hours of Sleep", title = "Histogram and Density Plot of Daily Sleep Hours with Median Line") +
  theme_minimal()

```

## 3. **Sleep Trends Over Time:** Create a time series plot of daily sleep duration to examine any trends or seasonality across the year.
```{r sleep_trends, TRUE}
ggplot(sleep_daily_clean, aes(x = date, y = sleep_hours)) +
  geom_line(color = "skyblue") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 5), color = "red") +
  labs(x = "Date", y = "Hours of Sleep", title = "Daily Sleep Duration Over Time") +
  theme_minimal()

ggplot(sleep_daily_clean %>% filter(date > "2024-01-01"), aes(x = date, y = sleep_hours)) +
  geom_line(color = "skyblue") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 5), color = "red") +
  labs(x = "Date (Jan-Sept 2024)", y = "Hours of Sleep", title = "Daily Sleep Duration Over Time") +
  theme_minimal()
```


## 4. **Monthly Averages:** Calculate and plot the average sleep duration per month to observe any seasonal variations.

```{r monthly_averages, TRUE}
sleep_monthly <- sleep_daily_clean %>%
  group_by(month) %>%
  summarise(median_sleep_hours = round(median(sleep_hours, na.rm = TRUE), 2))
sleep_monthly

#add text indicating median on top of each bar
#increase range of y axis to 10
ggplot(sleep_monthly, aes(x = month, y = median_sleep_hours)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  geom_text(aes(label = median_sleep_hours), vjust = -0.5, color = "black") +
  labs(x = "Month", y = "Median Sleep Hours", title = "Monthly Median Sleep Duration") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
  ) +
  ylim(0, 10)

#make a christmas tree plot instead of bar chart

ggplot(slee)

```
## 5. **Weekday vs. Weekend Sleep:** 
Compare sleep duration between weekdays and weekends using boxplots or bar plots.
```{r weekday_vs_weekend, TRUE}
sleep_daily_clean <- sleep_daily_clean %>%
  mutate(weekend = ifelse(week_day %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))

ggplot(sleep_daily_clean, aes(x = weekend, y = sleep_hours, fill = weekend)) +
  geom_boxplot() +
  labs(x = "Day Type", y = "Hours of Sleep", title = "Sleep Duration Comparison: Weekday vs. Weekend") +
  theme_minimal()

#plot the distribution of sleep hours for weekdays and weekends
ggplot(sleep_daily_clean, aes(x = sleep_hours, fill = weekend)) +
  geom_histogram(aes(y = ..density..), binwidth = 1, alpha = 0.7, position = "identity") +
  geom_density(alpha = 0.3) +
  labs(x = "Hours of Sleep", title = "Sleep Duration Distribution: Weekday vs. Weekend") +
  theme_minimal()

```



You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
